\section{Algorithm: SC-DenseAttrib}
\label{sec:algorithm}

\Cref{alg:sc-dense} presents the complete SC-DenseAttrib algorithm.
The procedure consists of five stages: support series computation, donor pool construction, weight estimation, causal effect estimation, and placebo-based inference.

\begin{algorithm}[t]
\caption{SC-DenseAttrib: Synthetic Control for Dense Pattern Attribution}
\label{alg:sc-dense}
\begin{algorithmic}[1]
\REQUIRE Transaction sequence $\mathcal{D}$, treated pattern $P^*$, candidate patterns $\mathcal{P}$, intervention time $t_0$, window size $W$
\ENSURE Causal effect $\hat{\tau}$, p-value $p$
\STATE Compute support series $\mathbf{s}_{P^*}$ via \Cref{def:windowed-support}
\STATE Build donor pool $\mathcal{D}(P^*) \leftarrow \{P \in \mathcal{P} : P \cap P^* = \emptyset\}$
\FOR{each $P_j \in \mathcal{D}(P^*)$}
    \STATE Compute $\mathbf{s}_{P_j}$ and filter by variance
\ENDFOR
\STATE Form donor matrix $\mathbf{D} \in \mathbb{R}^{T \times J}$
\STATE Estimate weights $\mathbf{w}^* \leftarrow \arg\min_{\mathbf{w} \in \Delta^J} \|\mathbf{s}_{P^*}^{\mathrm{pre}} - \mathbf{D}^{\mathrm{pre}} \mathbf{w}\|^2$
\STATE Compute counterfactual $\hat{\mathbf{s}}^{\mathrm{CF}} \leftarrow \mathbf{D} \mathbf{w}^*$
\STATE Compute $\hat{\tau}(t) \leftarrow \sigma_{P^*}(t) - \hat{\sigma}^{\mathrm{CF}}(t)$ for all $t$
\STATE $r^* \leftarrow \mathrm{RMSPE}_{\mathrm{post}}(\hat{\tau}) / \mathrm{RMSPE}_{\mathrm{pre}}(\hat{\tau})$
\FOR{each donor $P_j$ as placebo treated unit}
    \STATE Compute placebo effect and ratio $r_j$
\ENDFOR
\STATE $p \leftarrow |\{j : r_j \geq r^*\}| / (J + 1)$
\RETURN $\hat{\tau}$, $p$
\end{algorithmic}
\end{algorithm}

\subsection{Weight Estimation}

The optimization in~\eqref{eq:weight-opt} is a convex quadratic program over the simplex.
We solve it using the Frank-Wolfe (conditional gradient) algorithm, which iteratively moves toward the minimizing vertex of the simplex.
Each iteration requires $O(J)$ for the gradient computation and $O(J)$ for the vertex identification, yielding an overall complexity of $O(KJ)$ per weight estimation, where $K$ is the number of iterations (typically $K \leq 1000$).

\subsection{Donor Pool Construction}

The item-disjointness criterion (\Cref{def:donor-pool}) ensures that control patterns are not directly affected by interventions targeting items in $P^*$.
We additionally filter donors by requiring minimum variance in the pre-intervention period (to avoid degenerate flat trajectories) and optionally by pre-intervention correlation with the treated series.

\subsection{Complexity Analysis}

\begin{theorem}[Computational Complexity]
\label{thm:complexity}
SC-DenseAttrib runs in $O(N \cdot |\mathcal{P}| \cdot |P_{\max}| + J^2 \cdot t_0 + J \cdot (J^2 \cdot t_0))$ time, where $|P_{\max}|$ is the maximum pattern length, $J = |\mathcal{D}(P^*)|$ is the donor pool size, and $t_0$ is the pre-intervention period length.
\end{theorem}

\begin{proof}
Support series computation takes $O(N \cdot |P|)$ per pattern.
Donor pool construction is $O(|\mathcal{P}| \cdot |P^*|)$ for set intersection checks.
Weight estimation is $O(J^2 \cdot t_0)$ for the quadratic form.
Placebo tests repeat weight estimation $J$ times, each with $J$ donors, giving $O(J \cdot J^2 \cdot t_0)$ total.
\end{proof}
