\section{Introduction}
\label{introduction}

Recent increases in the volume of transaction data have rendered data analysis indispensable across diverse fields. The dynamic nature of data patterns, which often change over time, necessitates methods that can capture local variations. Transactions occurring with high frequency during specific intervals may reflect underlying factors such as seasonality, promotional activities, or societal trends. Detecting these dense patterns enables the extraction of significant insights, thereby facilitating effective decision-making and strategic planning.

Extracting dense patterns is challenging because it requires assessing the local frequency variations of itemsets rather than relying solely on global frequency metrics. Although numerous methods for itemset extraction exist \cite{agrawal1994fast}\cite{zaki1997new}\cite{han2000mining}, they primarily focus on overall frequency and fail to capture local fluctuations. Methods that consider the gaps between itemset occurrences have been proposed \cite{tanbeer2009discovering}\cite{kiran2017discovering}\cite{mahanta2005finding}\cite{fournier2021mining}, but our preliminary experiments revealed that their extraction accuracy is highly sensitive to the maximum allowed gap parameter. Moreover, these approaches do not exhaustively extract all dense patterns.

To address these issues, we propose two novel methods. The first method, termed \emph{Apriori-window}, is an exact solution that leverages a sliding window mechanism to evaluate local frequency variations without the need for a fixed maximum gap parameter. While this approach guarantees high extraction accuracy, its computational cost increases when itemsets occur frequently due to the exhaustive scanning of all occurrence points.

The second method, referred to as the \emph{Plant Model}, is an approximate solution that implements a swarm intelligence algorithm to detect dense intervals. By limiting scanning to these regions, the Plant Model reduces computational overhead while maintaining a robust performance. Together, these methods overcome the limitations of existing approaches by mitigating parameter sensitivity and reducing computational costs, thereby enabling efficient and accurate extraction of dense patterns from transaction data.

In Section 2 of this paper, we review related research on frequent itemset mining, variable frequency extraction, and swarm intelligence applications. Section 3 defines the problem. In Section 4, we present the proposed methods, \emph{Apriori-window} and the \emph{Plant Model}. Section 5 details the experimental setup, results, and discussion. We conclude in Section 6 with a brief summary and mention of future research directions.
