\documentclass[conference]{IEEEtran}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{algorithm}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{xcolor}

\begin{document}

\title{Anti-Dense Intervals and Contrast Dense Patterns:\\
Symmetric Extensions of Dense Interval Mining}

\author{
\IEEEauthorblockN{Anonymous Author(s)}
}

\maketitle

\begin{abstract}
Dense interval mining identifies contiguous time periods where a
pattern's sliding-window support exceeds a threshold, effectively
capturing epochs of sustained high activity in transactional data.
However, the existing framework is asymmetric: it characterizes
pattern \emph{presence} but provides no mechanism for characterizing
pattern \emph{absence} or for comparing temporal structures across
regimes.
In this paper, we propose two symmetric extensions.
First, we define \emph{Anti-Dense Intervals}---maximal contiguous
periods where support remains below a low threshold---and prove a
reverse monotonicity property that enables efficient mining.
Second, we introduce \emph{Contrast Dense Patterns}, a framework for
comparing dense interval structures across two temporal regimes.
We define a complete taxonomy of four topology change types
(Emergence, Vanishing, Amplification, Contraction) and provide a
permutation-based statistical test with false discovery rate control.
Both extensions share the same $O(n \log n)$ time complexity as
standard dense interval detection.
Experiments on synthetic data demonstrate anti-dense interval
recovery with F1 = 0.789, topology change classification with 76\%
accuracy, and statistically significant detection of
campaign-induced pattern vanishing ($p = 0.005$).
Our framework enables new analytical capabilities including the
identification of pattern disappearance epochs and the attribution
of structural changes to external interventions.
\end{abstract}

\begin{IEEEkeywords}
dense interval mining, anti-dense intervals, contrast patterns,
temporal pattern analysis, emerging patterns, permutation test
\end{IEEEkeywords}

\input{sec/01_introduction}
\input{sec/02_related_work}
\input{sec/03_problem_definition}
\input{sec/04_proposed_method}
\input{sec/05_experiments}
\input{sec/06_conclusion}

\bibliographystyle{IEEEtran}
\bibliography{refs}

\end{document}
