% ==============================================================
\section{Related Work}\label{sec:related}
% ==============================================================

Our work spans several research areas.
We review each and highlight the gap that our contributions address.

\subsection{Frequent and Dense Interval Mining}

The Apriori algorithm~\cite{agrawal1994apriori} and
FP-Growth~\cite{han2000fpgrowth} established the foundations of
frequent itemset mining.
Dense interval mining extends this framework by identifying
contiguous temporal intervals where a pattern's sliding-window
support exceeds a threshold.
While dense intervals effectively capture periods of high activity,
the framework does not provide a symmetric mechanism for
characterizing periods of sustained \emph{low} activity.
Our anti-dense interval concept fills this gap.

\subsection{Rare and Infrequent Pattern Mining}

Mining infrequent or rare patterns has received growing attention.
RP-Tree~\cite{szathmary2011rptree} extends FP-Growth to mine rare
itemsets.
Haglin and Manning~\cite{haglin2007minimal} introduced minimal
infrequent itemsets.
ERIM~\cite{akdas2024erim} combines four rare itemset mining
algorithms in an ensemble.
Recent work on rare partial periodic patterns in data
streams~\cite{kiran2025r3pstream} addresses temporal rarity in the
streaming setting.
Negative association rules~\cite{wu2024negative} capture item
co-absence but in a static, non-temporal framework.

These approaches share a focus on \emph{globally} infrequent
patterns but do not characterize \emph{temporal intervals} of
infrequency.
A pattern may be globally frequent yet exhibit sustained periods
of absence---a phenomenon that our anti-dense intervals capture
directly.

\subsection{Emerging and Contrast Patterns}

Dong and Li~\cite{dong1999emerging} introduced emerging patterns
(EPs)---itemsets whose support ratio between two datasets exceeds a
threshold.
The contrast data mining framework~\cite{dong2012contrast} generalizes
this to multi-class comparisons.
Garc\'ia-Vico et al.~\cite{garciavico2018ep} provide a comprehensive
taxonomy of EP mining approaches.
Loyola-Gonz\'alez et al.~\cite{loyola2020contrast} survey contrast
pattern mining with statistical validation.
Recent work by Li et al.~\cite{li2025hucspm} extends contrast
patterns to high-utility sequential patterns.

Emerging patterns compare \emph{support values} between datasets.
Our contrast dense patterns compare \emph{dense interval structures}
between temporal regimes---a fundamentally different and richer
comparison that captures not just whether support changed, but
\emph{how the temporal pattern of support changed}.

\subsection{Change Point Detection}

Change point detection identifies time points where the statistical
properties of a time series change.
CUSUM~\cite{page1954cusum} detects mean shifts sequentially.
PELT~\cite{killick2012pelt} provides optimal multiple change point
detection in $O(n)$ time.
Bai and Perron~\cite{bai2003structural} address multiple structural
breaks in regression models.
Truong et al.~\cite{truong2020ruptures} provide a comprehensive
Python library for offline change point detection.
Recent applications include financial time series
analysis~\cite{chen2025pelt}.

These methods operate on \emph{univariate or multivariate time series}
and detect general distributional changes.
They do not consider the specific structure of \emph{pattern-level}
dense intervals or classify the \emph{type} of structural change
(emergence, vanishing, etc.) as our framework does.

\subsection{Concept Drift Detection}

Concept drift detection identifies changes in the data-generating
process over time.
ADWIN~\cite{bifet2007adwin} uses adaptive windowing with statistical
bounds.
DDM~\cite{gama2004ddm} detects drift through error rate monitoring.
ADWIN-U~\cite{assis2024adwinu} extends ADWIN to unsupervised settings.
Cerqueira et al.~\cite{cerqueira2024benchmark} benchmark fully
unsupervised drift detectors.

Concept drift detection signals that ``something changed'' but does
not provide a structured characterization of \emph{what} changed at
the level of individual patterns.
Our topology change taxonomy (Emergence, Vanishing, Amplification,
Contraction) provides exactly this interpretable characterization.

\subsection{Temporal and Periodic Pattern Mining}

Temporal pattern mining has been studied extensively.
3P-ECLAT~\cite{3peclat2024} mines partial periodic patterns in
columnar databases.
TaTIRP~\cite{tatirp2025} discovers targeted time-interval related
patterns using Allen's temporal relations.

These approaches focus on periodicity or temporal relations between
events, which is complementary to our focus on dense/anti-dense
intervals defined by sliding-window support thresholds.
