% ==============================================================
\section{Conclusion}\label{sec:conclusion}
% ==============================================================

We have presented two symmetric extensions of the dense interval
mining framework: \emph{Anti-Dense Intervals} and \emph{Contrast
Dense Patterns}.

Anti-dense intervals formalize the dual concept of dense intervals,
capturing sustained periods of pattern \emph{absence}.
We proved a reverse monotonicity property that enables efficient
top-down search strategies and showed that anti-dense intervals
can be computed with the same time complexity as dense intervals.

Contrast dense patterns provide a structured framework for comparing
dense interval configurations across temporal regimes.
Our taxonomy of four topology change types (Emergence, Vanishing,
Amplification, Contraction) gives interpretable characterizations
of how patterns evolve over time.
The accompanying permutation test, combined with Benjamini--Hochberg
correction, ensures statistical rigor.

Experimental evaluation on synthetic data demonstrated effective
anti-dense interval recovery (F1 = 0.789) and reliable topology
change classification (76\% accuracy).
Application to simulated campaign termination scenarios showed that
the method can identify campaign-dependent patterns with high
statistical significance ($p = 0.005$), while correctly classifying
unaffected patterns as stable.

\paragraph{Limitations.}
The current framework requires a pre-specified regime boundary $\tau$.
Automatic boundary detection via change point methods could extend
applicability.
The permutation test assumes exchangeability of timestamps under the
null hypothesis, which may be violated in the presence of strong
temporal autocorrelation.

\paragraph{Future Work.}
Several directions merit further investigation:
(i) integration with automatic change point detection (e.g.,
PELT~\cite{killick2012pelt}) to discover regime boundaries from
the data;
(ii) extension to multi-regime comparisons beyond the two-regime
case;
(iii) application to real-world retail datasets (Dunnhumby, Online
Retail) with known campaign calendars;
(iv) development of more efficient pruning strategies exploiting
the reverse monotonicity of anti-dense intervals;
and (v) exploration of dense-to-anti-dense transition dynamics as
a characterization of pattern lifecycle.
