% ==============================================================
\section{Introduction}\label{sec:intro}
% ==============================================================

Frequent itemset mining~\cite{agrawal1994apriori} is a foundational technique in data mining, with applications ranging from market basket analysis to bioinformatics.
The Apriori algorithm and its successors exploit the \emph{anti-monotonicity} (downward closure) property of support: if an itemset is infrequent, all its supersets must also be infrequent.
This property enables aggressive pruning of the combinatorial search space, making large-scale mining tractable.

However, anti-monotonicity is a double-edged sword.
By design, it eliminates all patterns whose global support falls below the minimum threshold.
In many real-world scenarios, the most interesting patterns are precisely those that are \emph{rare}---occurring too infrequently to survive the global support filter---yet exhibit meaningful temporal structure.
Consider the following motivating examples:

\begin{itemize}
\item \textbf{Flash sales in retail}: A promotional bundle of luxury items may be purchased by only 0.1\% of customers overall, but during a two-week flash sale, these items co-occur in 30\% of daily transactions. Standard frequent pattern mining with a 1\% threshold would never discover this pattern.

\item \textbf{Security incidents}: A specific combination of network events (e.g., port scanning + credential stuffing + lateral movement) may occur in fewer than 0.01\% of all log entries, but during an active attack, these events cluster within minutes. Anomaly detection methods flag individual events but cannot identify the multi-item co-occurrence pattern.

\item \textbf{Seasonal co-purchases}: Certain product combinations (e.g., sunscreen + insect repellent + camping gear) may have very low annual support but spike during specific weeks. Understanding these temporal bursts is valuable for inventory planning.
\end{itemize}

These examples share a common structure: the patterns are \emph{globally rare} (low overall support) but \emph{locally dense} (high support within a narrow temporal window).
We call such patterns \emph{Rare Dense Patterns} (RDPs).

Existing approaches address parts of this problem but not the whole:

\begin{enumerate}
\item \textbf{Rare pattern mining} algorithms (RP-Tree~\cite{tsang2011rptree}, Apriori-Rare~\cite{szathmary2007rare}, MFI mining~\cite{haglin2007mfi}) discover globally infrequent itemsets but are agnostic to temporal structure. They cannot distinguish between a uniformly rare pattern (noise) and a temporally concentrated rare pattern (signal).

\item \textbf{Burst detection} (Kleinberg~\cite{kleinberg2003bursty}) identifies temporal concentration in univariate event streams but does not handle the combinatorial structure of multi-item co-occurrence patterns.

\item \textbf{Anomaly detection} (Isolation Forest~\cite{liu2008iforest}, LOF~\cite{breunig2000lof}) operates at the point or transaction level and cannot identify structured itemset patterns.

\item \textbf{Sliding-window frequent pattern mining} discovers locally frequent patterns but does not filter for global rarity, producing outputs dominated by globally frequent patterns.
\end{enumerate}

In this paper, we make the following contributions:

\begin{enumerate}
\item We formalize the \emph{Rare Dense Pattern} (RDP) mining problem, defining global rarity and local density conditions, and characterize the class of patterns that satisfy both (\S\ref{sec:problem}).

\item We prove a \emph{Weak Anti-Monotonicity} theorem: while the combined RDP condition lacks classical anti-monotonicity, the local density component is anti-monotone, enabling Apriori-style pruning (\S\ref{sec:problem}).

\item We propose a \emph{Two-Phase Mining} algorithm that is provably sound and complete---it discovers exactly the set of all RDPs (\S\ref{sec:algorithm}).

\item We conduct comprehensive experiments on synthetic data showing 100\% recovery rate, 80\%+ false positive reduction via the rarity filter, and near-linear scalability to 100K transactions (\S\ref{sec:experiments}).
\end{enumerate}

The remainder of this paper is organized as follows.
\S\ref{sec:related} reviews related work.
\S\ref{sec:problem} formalizes the problem and states our theoretical results.
\S\ref{sec:algorithm} presents the Two-Phase Mining algorithm.
\S\ref{sec:experiments} reports experimental results.
\S\ref{sec:conclusion} concludes with future directions.
