% ==============================================================
\section{Conclusion}\label{sec:conclusion}
% ==============================================================

We introduced the \emph{Rare Dense Pattern} (RDP) mining problem, a novel formulation that targets itemsets satisfying two simultaneous conditions: global rarity (low overall support) and local temporal density (high co-occurrence count within at least one sliding window).
This class of patterns captures practically important phenomena---promotional bursts, security incidents, seasonal co-purchases---that are invisible to both standard frequent pattern mining (which prunes rare patterns) and standard rare pattern mining (which ignores temporal structure).

Our key theoretical contribution is the \emph{Weak Anti-Monotonicity} property: while the combined RDP condition is neither monotone nor anti-monotone, the local density component alone is anti-monotone.
This enables Apriori-style pruning in a \emph{Two-Phase Mining} algorithm that first discovers all locally dense patterns, then filters by global rarity.
We proved that this algorithm is sound and complete---it discovers exactly the set of all RDPs.

Experimental results on synthetic data demonstrate 100\% recovery of ground-truth RDPs, 80\%+ reduction in false positives via the rarity filter, and near-linear scalability to 100K transactions.

\paragraph{Limitations.}
The current algorithm assumes a fixed window size $W$ and density threshold $\theta$.
Automatically selecting these parameters based on data characteristics remains an open problem.
The global rarity condition uses a single threshold $\sigma_{\max}$; per-item or per-itemset-size thresholds could improve flexibility.

\paragraph{Future Work.}
Several directions merit investigation:
\begin{enumerate}
\item \textbf{Multi-scale analysis}: Detecting RDPs across multiple window sizes simultaneously, potentially using hierarchical decomposition similar to Kleinberg's multi-level burst detection~\cite{kleinberg2003bursty}.

\item \textbf{Streaming setting}: Adapting the algorithm for online mining where transactions arrive continuously, maintaining RDPs incrementally.

\item \textbf{Rarity-aware pruning}: Developing tighter pruning strategies that exploit the rarity condition during Phase~1, potentially reducing the Phase~1 candidate set.

\item \textbf{Event attribution}: Linking discovered RDPs to external events (promotions, incidents) to explain \emph{why} a rare pattern became locally dense at a specific time.

\item \textbf{Large-scale evaluation}: Applying the framework to real-world datasets (retail, network logs, healthcare) to validate the practical utility of RDPs.
\end{enumerate}
