% ==============================================================
% Event Attribution of Frequent Pattern Dynamics
% ==============================================================
% Compilation:
%   pdflatex main.tex && bibtex main && pdflatex main.tex && pdflatex main.tex
% ==============================================================
\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

% --- Standard packages ---
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{amsthm}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{threeparttable}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{url}
\usepackage{multirow}

% --- Figure path ---
\graphicspath{{../manuscript/fig/}}

% --- Theorem environments ---
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{claim}{Claim}
\newtheorem{lemma}{Lemma}
\newtheorem{remark}{Remark}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Event Attribution of Frequent Pattern Dynamics:\\Permutation Testing with FDR Control}

\author{
  \IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
  \IEEEauthorblockA{
    \textit{Dept. Name of Organization}\\
    \textit{Name of Organization}\\
    City, Country\\
    email@example.com}
  \and
  \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
  \IEEEauthorblockA{
    \textit{Dept. Name of Organization}\\
    \textit{Name of Organization}\\
    City, Country\\
    email@example.com}
}

\maketitle
\pagestyle{plain}
\pagenumbering{arabic}

\begin{abstract}
Frequent pattern mining discovers co-occurrence frequency of itemsets but lacks
the ability to explain which external events (promotions, holidays, policy changes, etc.)
drive temporal changes in pattern support.
This paper formalizes this task as the \textbf{event attribution problem}
and identifies its structural challenges as a multiple testing problem with FDR control:
correlated false positives, temporal confounding, and hypothesis space scale.
We propose a five-step pipeline comprising change point extraction, attribution scoring,
circular-shift permutation testing, Benjamini--Hochberg FDR correction,
and Union-Find deduplication.
On synthetic data, we confirm FDR~0.070 ($\leq \alpha = 0.10$) under null conditions
and demonstrate the necessity of each pipeline component through ablation analysis.
Comparison against four related methods (Wilcoxon, CausalImpact, ITS, EventStudy)
achieves best-or-equal F1 across all experimental conditions.
Exploratory application to real-world retail data (Dunnhumby, 276K baskets)
confirms 100\% campaign coverage with an average of 3.0 attributions per campaign.
\end{abstract}

\begin{IEEEkeywords}
frequent pattern mining, event attribution, change point detection,
permutation test, multiple testing correction, support time series
\end{IEEEkeywords}

\input{sec/01_intro}
\input{sec/02_related_work}
\input{sec/03_problem_definition}
\input{sec/04_proposed_method}
\input{sec/05_experiments}
\input{sec/06_conclusion}

% --- Acknowledgment ---
\section*{Acknowledgment}
The authors acknowledge that portions of the text in
this paper were generated with Anthropic's large language
model Claude~\cite{claude2025}.

% --- References ---
\newpage
\bibliographystyle{IEEEtran}
\bibliography{refs}

% --- Appendices ---
\clearpage
\appendices
\input{sec/07_appendix}

\end{document}
